Experience Report: Haskell in Computational Biology

### 2012-03-06 mrfy_experience_report.tex
newlines after sentences
intro section receives the Ramsey treatment

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more. The function of a protein is determined by its structure, 
and this structure in turn is determined by the specific sequence of the twenty 
different amino acids that make up the protein, but proteins with significantly 
different amino acid sequences may share similar structure and function. This 
amino acid sequence is ultimately determined by the DNA sequence we call a 
gene. Given the genetic sequence of a protein, biologists wish to know its 
specific cellular function. The best path to this is to find other proteins of 
similar structure, which likely share similar function. This problem is called 
\textit{homology detection}.

Computational biologists address the homology detection problem by building 
algorithms to find similar proteins given an amino acid sequence. When there 
are nearly-identical sequences for proteins of known structure, this is a 
simple string-matching problem. When there are no identical sequences, but 
still some reasonably similar sequences, one family of algorithms called 
\textit{hidden Markov models} excels at homology detection. We will explain an 
algorithm for detecting reasonably similar sequences, and then move on to 
explain MRFy, our novel approach for detecting more distantly related sequences 
for proteins that share similar structure and function.

@@

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position. Each edge in the 
graph has a weight which is the probability of the associated state transition. 
Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
which correspond to columns of the original alignment. What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

@@

The Viterbi algorithm comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements. The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type. 
Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
or \texttt{consNoPath} as a parameter, the path would or would not be 
allocated. We found this higher-order function approach to be simple in code 
and in concept, and it imposes negligible run-time overhead as compared to 
having two distinct functions.

@@

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{matt}. 
Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
data structures such as mutable oct-trees, vectors, and arrays, with 
significant amounts of clever pointer arithmetic. Our enhancement involved 
modifying Matt to incorporate sequence information along with the structural 
aligner. While we initially estimated this would be a three-month project, it 
took most of a year because the mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: since 
invariants were not clear, nearly every change resulted in new segfaults. One 
enhancement, partial alignments (essentially a cosmetic manipulation of the 
output, but requiring deep information about some of the data structures) was 
deemed infeasible. We believe we could have completed a ground-up rewrite in 
Haskell, with most of the run-time performance, in fewer than nine months.

@@

Haskell makes it difficult to write code in haste, because Haskell militates 
towards careful type-checking and clearly thought-out function contracts. 
Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
programmer who attempts to write code quickly with the goal of later making it 
work correctly. We have found MRFy to be easier to enhance and maintain than 
our existing C++ code bases. Despite the learning curve of Haskell, we 
implemented most of MRFy in roughly three months of part-time work.

@@

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

>>>>

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
This problem is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name. ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm, can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

\emph{[We need an argument here saying why an understanding of the
    algorithms is relevant to the claims made in the experience
    report, then say how we present the algorithms, then say what
    experiential conclusions this presentation supports.]}

@@

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}).
Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position.
Each edge in the 
graph has a weight which is the probability of the associated state transition.

Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
which correspond to columns of the original alignment.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

@@

The Viterbi algorithm comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model).

To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence.
In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.
The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type.

Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
or \texttt{consNoPath} as a parameter, the path would or would not be 
allocated.
We found this higher-order function approach to be simple in code 
and in concept, and it imposes negligible run-time overhead as compared to 
having two distinct functions.

@@

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{matt}.

Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
data structures such as mutable oct-trees, vectors, and arrays, with 
significant amounts of clever pointer arithmetic.
Our enhancement involved 
modifying Matt to incorporate sequence information along with the structural 
aligner.
While we initially estimated this would be a three-month project, it 
took most of a year because the mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: since 
invariants were not clear, nearly every change resulted in new segfaults.
One 
enhancement, partial alignments (essentially a cosmetic manipulation of the 
output, but requiring deep information about some of the data structures) was 
deemed infeasible.
We believe we could have completed a ground-up rewrite in 
Haskell, with most of the run-time performance, in fewer than nine months.

@@

Haskell makes it difficult to write code in haste, because Haskell militates 
towards careful type-checking and clearly thought-out function contracts.

Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
programmer who attempts to write code quickly with the goal of later making it 
work correctly.
We have found MRFy to be easier to enhance and maintain than 
our existing C++ code bases.
Despite the learning curve of Haskell, we 
implemented most of MRFy in roughly three months of part-time work.

@@

\bibliographystyle{plainnat}
